---
title: "Coursera Capstone Project (Case 1)"
output:
  html_document:
    df_print: paged
---

### Cyclistic_Exercise_Full_Year_Analysis

Cyclistic, a bike-share company seeks to understand similar and differential trends between its annual members and casual riders for conversion of casual customers to members through targeted marketing strategies.
Stakeholders include member and casual customers, the government of Chicago and Cyclistic.

```{r Install and load libraries, echo=TRUE, message=FALSE, warning=FALSE}
# Install and load required packages
library(tidyverse)  # for data import, wrangling and visualization

install.packages("geosphere")
library(geosphere)  # for calculating distances from latitudes and longitudes

install.packages("scales")
library(scales)  # to aid scaling of plot axes
```

```{r Set working directory, message=FALSE, warning=FALSE}

# set working directory
setwd("C:/Users/abayomia/Downloads/coursera/capstone_project/case_one_cyclistic/updated_files")

# confirm working directory
getwd()
```

The year 2022 bike-share data was sourced from a storage bucket at cyclic 2022 bikeshare data on Amazon cloud services.
https://divvy-tripdata.s3.amazonaws.com/index.html

### STEP 1: COLLECT DATA
```{r Step 1: Collect Data, message=FALSE, warning=FALSE}

# Upload Cyclistic datasets (csv files) here
jan <- read_csv("202201-divvy-tripdata.csv")
feb <- read_csv("202202-divvy-tripdata.csv")
mar <- read_csv("202203-divvy-tripdata.csv")
apr <- read_csv("202204-divvy-tripdata.csv")
may <- read_csv("202205-divvy-tripdata.csv")
jun <- read_csv("202206-divvy-tripdata.csv")
jul <- read_csv("202207-divvy-tripdata.csv")
aug <- read_csv("202208-divvy-tripdata.csv")
sep <- read_csv("202209-divvy-tripdata.csv")
oct <- read_csv("202210-divvy-tripdata.csv")
nov <- read_csv("202211-divvy-tripdata.csv")
dec <- read_csv("202212-divvy-tripdata.csv")
```

### STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
```{r Step 2: Wrangle Data and Combine into a Single File}

# Compare column names in each of the files to check for dispcrepancies
colnames(jan) == colnames(feb)
colnames(jan) == colnames(mar)
colnames(jan) == colnames(apr)
colnames(jan) == colnames(may)
colnames(jan) == colnames(jun)
colnames(jan) == colnames(jul)
colnames(jan) == colnames(aug)
colnames(jan) == colnames(sep)
colnames(jan) == colnames(oct)
colnames(jan) == colnames(nov)
colnames(jan) == colnames(dec)

```
Column names are consistent throughout tables

```{r Step 2.1: Stack monthly data frames into one big data frame}

# combine all months into a single dataset
all_trips <- bind_rows(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec)

# preview new dataset
glimpse(all_trips)
```

### STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
```{r Step 3: Inspect the new table that has been created}

# show column names of table
colnames(all_trips)

# show the number of rows of data frame
nrow(all_trips)

# show dimensions of data frame, number of rows and columns
dim(all_trips)

# show the first 6 rows of the data frame
head(all_trips)

# show the last 6 rows of data frame
tail(all_trips)

# compactly display structure of data frame
str(all_trips)

# show concise statistical summary of each variable or column, mainly for numerics
summary(all_trips)

# get a glimpse of data using dplyr package
glimpse(all_trips)
```
There are 5,667,717 rows (observations) and 15 columns (variables) in "all_trips" table

There are a few issues that need to be fixed:

1. "started_at" and "ended_at" columns are in character format and would need to be changed to datetime or POSIXct.

2. "ride_length" is in "hms difftime" time format and should be changed to numerical format for easy plotting and computation.

```{r Step 3.1: Convert to appropriate formats}

# create a copy of the data for possible reverts
df <- all_trips

# use lubridate package to convert "started_at" and "ended_at" character formats to datetime
df$started_at <- dmy_hm(df$started_at)
df$ended_at <- dmy_hm(df$ended_at)

# check to confirm successful conversion
class(df$started_at[2])
class(df$ended_at[2])

# convert "ride_length" to numerical format
df$ride_length <- as.numeric(df$ride_length)

# check to note successful conversion
is.numeric(df$ride_length)

# add columns specifying hour, day, day of week and month and make copy
df <- df %>% 
  mutate(hour = hour(started_at), day_of_week = wday(started_at, label = TRUE), month = month(started_at, label = TRUE))

# confirm column count
dim(df)

# use geosphere to calculate distance from latitude and longitude data
df$distance <- distGeo(df[,c("start_lng", "start_lat")], df[,c("end_lng", "end_lat")]) %>% round(digits=2)

# confirm column count
dim(df)
```

```{r Step 3.2: Remove bad data}

# remove observations where both ride_length and distance are less than or equal to zero, based on the assumption of maintenance repairs
df <- df %>% 
  filter(!(ride_length <= 0 & distance <= 0))

# remove observations with null values in "ride_length" column
df <- df %>% 
  drop_na(ride_length)
```

### STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
```{r Step 4: Descriptive analysis on ride_length (all figures in seconds)}

mean(df$ride_length)
median(df$ride_length)
max(df$ride_length)
min(df$ride_length)

# compare members and casual users
aggregate(df$ride_length ~ df$member_casual, FUN = mean)
aggregate(df$ride_length ~ df$member_casual, FUN = median)
aggregate(df$ride_length ~ df$member_casual, FUN = max)
aggregate(df$ride_length ~ df$member_casual, FUN = min)

# averge ride time by day of week for members vs casual riders
aggregate(df$ride_length ~ df$member_casual + df$day_of_week, FUN = mean)

aggregate(df$ride_length ~ df$member_casual + df$day_of_week, FUN = max)

aggregate(df$ride_length ~ df$member_casual + df$day_of_week, FUN = min)

# averge ride time by month for members vs casual riders
aggregate(df$ride_length ~ df$member_casual + df$month, FUN = mean)

aggregate(df$ride_length ~ df$member_casual + df$month, FUN = max)

aggregate(df$ride_length ~ df$member_casual + df$month, FUN = min)
```
```{r}
colnames(df)
```

#### STEP 4.1: PLOT COUNT OF RIDES 
```{r Step 4.1.1: Group and visualize number of rides by hour and rider type}

# visualize number of rides by hour and rider type in column chart
df %>% 
  group_by(member_casual, hour) %>% 
  summarize(number_of_rides = n()
            , average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = hour, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  scale_x_continuous(breaks = scales::pretty_breaks(n = 23)) +
  labs(title = "Number of Rides by Hour"
       , subtitle = "Number of rides by rider type and hour"
       , fill = "Rider Type") +
  xlab("Time (Hour)") +
  ylab("Number of Rides")

# save column plot
ggsave("ride_count_by_hour_col.png", width = 5000, height = 3500, units = "px", dpi = 700)


# visualize number of rides by hour and rider type in line chart
df %>% 
  group_by(member_casual, hour) %>% 
  summarize(number_of_rides = n()
            , average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = hour, y = number_of_rides, color = member_casual)) +
  geom_line(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  scale_x_continuous(breaks = scales::pretty_breaks(n = 23)) +
  labs(title = "Number of Rides by Hour"
       , subtitle = "Number of rides by rider type and hour"
       , color = "Rider Type") +  # fill is replaced with color
  xlab("Time (Hour)") +
  ylab("Number of Rides")

# save line plot
ggsave("ride_count_by_hour_line.png", width = 5000, height = 3500, units = "px", dpi = 700)
```

```{r Step 4.1.2: Group and visualize number of rides by day of week and rider type}

# visualize number of rides by day of week and rider type
df %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  labs(title = "Number of Rides by Day of Week",
       subtitle = "Number of rides by rider type and day of week",
       fill = "Rider Type") +
  xlab("Day of Week") +
  ylab("Number of Rides")

# save plot
ggsave("ride_count_by_weekday.png", width = 5000, height = 3500, units = "px", dpi = 700)
```

```{r Step 4.1.3: Group and visualize number of rides by month and rider type}

# visualize number of rides by month and rider type
df %>% 
  group_by(member_casual, month) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  labs(title = "Number of Rides by Month",
       subtitle = "Number of rides by rider type and month",
       fill = "Rider Type") +
  xlab("Month") +
  ylab("Number of Rides")

# save plot
ggsave("ride_count_by_month.png", width = 5000, height = 3500, units = "px", dpi = 700)
```

```{r Step 4.1.4: Group and visualize number of rides by season and rider type}

# visualize number of rides by season and rider type

# specify seasons from months in a vector
spring <- c('Mar', 'Apr', 'May')
summer <- c('Jun', 'Jul', 'Aug')
fall <- c('Sep', 'Oct', 'Nov')
winter <- c('Dec', 'Jan', 'Feb')

# add seasons as a variable in dataset
df <- df %>%
  mutate(season = case_when(
    month %in% spring ~ 'spring',
    month %in% summer ~ 'summer',
    month %in% fall ~ 'fall',
    month %in% winter ~ 'winter'
  ))

# visualization
df %>% 
  group_by(member_casual, season) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = factor(season, level=c('spring', 'summer', 'fall', 'winter')), y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  labs(title = "Number of Rides by Season",
       subtitle = "Number of rides by rider type and season",
       fill = "Rider Type") +
  xlab("Season") +
  ylab("Number of Rides")

# save plot
ggsave("ride_count_by_season.png", width = 5000, height = 3500, units = "px", dpi = 700)
```

#### STEP 4.2: PLOT AVERAGE RIDE LENGTH 
```{r Step 4.2.1: Group and visualize average ride length by hour and rider type}

# visualize average ride length by hour and rider type in column format
df %>% 
  group_by(member_casual, hour) %>% 
  summarize(number_of_rides = n()
            , average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = hour, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  scale_x_continuous(breaks = scales::pretty_breaks(n = 23)) +
  labs(title = "Average Trip Duration by Hour"
       , subtitle = "Average ride length by rider type and hour"
       , fill = "Rider Type") +
  xlab("Time (Hour)") +
  ylab("Average Ride Length (seconds)")

# save plot
ggsave("average_trip_duration_by_hour_col.png", width = 5000, height = 3500, units = "px", dpi = 700)


# visualize average ride length by hour and rider type in line format
df %>% 
  group_by(member_casual, hour) %>% 
  summarize(number_of_rides = n()
            , average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = hour, y = average_duration, color = member_casual)) +
  geom_line(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  scale_x_continuous(breaks = scales::pretty_breaks(n = 23)) +
  labs(title = "Average Trip Duration by Hour"
       , subtitle = "Average ride length by rider type and hour"
       , color = "Rider Type") +
  xlab("Time (Hour)") +
  ylab("Average Ride Length (seconds)")

# save plot
ggsave("average_trip_duration_by_hour_line.png", width = 5000, height = 3500, units = "px", dpi = 700)
```


```{r Step 4.2.2: Group and visualize average ride length by day of week and rider type}

# visualize average ride length by day of week and rider type
df %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  labs(title = "Average Trip Duration by Day of Week",
       subtitle = "Average ride length by rider type and day of week",
       fill = "Rider Type") +
  xlab("Day of Week") +
  ylab("Average Ride Length (seconds)")

# save plot
ggsave("average_trip_duration_by_weekday.png", width = 5000, height = 3500, units = "px", dpi = 700)
```

```{r Step 4.2.3: Group and visualize average ride length by month and rider type}

# visualize average ride length by month and rider type
df %>% 
  group_by(member_casual, month) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  labs(title = "Average Trip Duration by Month",
       subtitle = "Average ride length by rider type and month",
       fill = "Rider Type") +
  xlab("Month") +
  ylab("Average Ride Length (seconds)")

# save plot
ggsave("average_trip_duration_by_month.png", width = 5000, height = 3500, units = "px", dpi = 700)
```

```{r Step 4.2.4: Group and visualize average ride length by season and rider type}

# visualize average ride length by season and rider type
df %>% 
  group_by(member_casual, season) %>% 
  summarize(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = factor(season, level=c('spring', 'summer', 'fall', 'winter')), y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # scale y-axis to disregard scientific notation
  labs(title = "Average Trip Duration by Season",
       subtitle = "Average ride length by rider type and season",
       fill = "Rider Type") +
  xlab("Season") +
  ylab("Average Ride Length (seconds)")

# save plot
ggsave("average_trip_duration_by_season.png", width = 5000, height = 3500, units = "px", dpi = 700)
```

```{r Step 4.3: Export summary table to csv for visualization}

# aggregate average ride length based on rider type and day of week
avg_ride_length_weekly <- aggregate(df$ride_length ~ df$member_casual + df$day_of_week, FUN = mean)
avg_ride_length_weekly

# aggregate average ride length based on rider type and hour
avg_ride_length_hourly <- aggregate(df$ride_length ~ df$member_casual + df$hour, FUN = mean)
avg_ride_length_hourly

# aggregate average ride length based on rider type and month
avg_ride_length_monthly <- aggregate(df$ride_length ~ df$member_casual + df$month, FUN = mean)
avg_ride_length_monthly

# aggregate average ride length based on rider type and season
avg_ride_length_seasonally <- aggregate(df$ride_length ~ df$member_casual + df$season, FUN = mean)
avg_ride_length_seasonally

write.csv(avg_ride_length_weekly, file='avg_ride_length_weekly.csv')
write.csv(avg_ride_length_hourly, file='avg_ride_length_hourly.csv')
write.csv(avg_ride_length_monthly, file='avg_ride_length_monthly.csv')
write.csv(avg_ride_length_seasonally, file='avg_ride_length_seasonally.csv')
```
